---
title: "Inferential Statistics with the GSS dataset"
author: "Mobin Piracha"
date: "11/29/2020"
output: rmarkdown::github_document
---

## Setup

### Load Packages

```{r load-packages, message = FALSE}
library(tidyverse)
library(statsr)
library(ggthemes)
```

### Load Data

```{r load-data, message=FALSE}
load("gss.Rdata")
```

* * *

## Part I: Data

The General Social Survey has been used to monitor change and studying the growing complexity of American society since 1972. It is used to analyze data on contemporary American society to explain trends in attitudes, behaviors and functioning of American society. The GSS surveys U.S households using proportional survey technique. We use this technique because the population is composed of various subgroups and therefore, the survey samples according to the different subgroups, that all add up to make up the sample. The sample is drawn using an area probability design that randomly selects respondents across the United States to take part in the survey. The sample includes in person and telephone interviews. Therefore, the GSS uses random sampling and therefore the results of the survey as generalizable. Since there is no random assignment, the survey results are not causal. Therefore, the survey is generalizable to the larger population but not causal. 

* * *

## Part II: Research question

#### Exploratory Data Analysis

#### Research Question 1: 

####  (i) How has religious representation in the sample changed from 1973 to 2012. Create a line graph to show change overtime, and create a stacked barplot to show the change in each decade from 1973 to 2012. 

#### (ii) How has attendance in religious services changed between 1973 to 2013. Create a bar plot for attendance in religious services in 1973 and another in 2013. What differences do you see? 

#### (iii) How has confidence in organized religion changed, show decade long change with a stacked bar chart. 

#### Research Question 2: 

#### (i) For simplicity, reclassify the degree variable into those Americans with bachelor's degree and those without a bachelor's degree. How has bachelor's degreee attainment changed in American from 1973 to 2012? Create a line graph and and a stacked bar chart to show this overtime change. 

#### (ii) Have the proportion of Americans with bachelor degrees changed? Has the gendar ratio for attainment of a bachelor's degree changed from 1972 to 2012? Visualize this data. Have proportions of bachelor's degree attainment changed by race? Visualize this data.

#### (iii) Is there a relationship between confidence in education and education levels. Explore both variables, visualize the relationship between these variables and also create a correlation matrix for years 1973, 2000 and 2012.


#### Statistical Inference

#### Research Question 1: 

(i) We find that the mean education in Australia in 1975 was 10.04 and the mean education for Canada in 1975 was 9.18.Conduct a hypothesis test for means, finding out if the mean education in the United States in 1973 is greater than the mean education in Australia, conduct simialr hypothesis test for US and Canada. Lastly find the confidence intervel for mean education level in the United States.

(ii) Is there convincing evidence that the United States has seen a change in its education levels from 1973 to 2012.

(iii) We would like to esimtate the difference between two independent means. Therefore, conduct a hypothesis test evaluating whether the education levels of male is different than the education level of female. Find inference for both 1973 and 2012.

Let's conduct one more test comparing more than two means, in this case an ANOVA. Let's find out whether education levels are different based on race. Let's conduct this for both year 1973 and year 2012. 

#### Research Question 2: 

(i) According to sources, 22.44% of Canadians and 30% Germans have bachelor's degrees. Conduct a hypothesis test for proportions and find if the proportion of Americans with bachelor's degrees is greater than Canadians and Germans in year 2012. 

(ii) Is there convincing evidence that United States has seen a change in its bachelor's degree attainment levels between 2002 and 2012?

(iii) Conduct a hyppthesis test of two sample proportions and find if there is a relationship between bachelor's degree attainment and party ID (for simplicity we will assume that there are only two parties: Democratic and Republican Party and delete all other observations). We will conduct a confidence intervel to estimate whether there is a difference between democrats with a bachelor's degree and republican with a bachelor's degree for years 1993 and 2012 

(iv) Conduct a chi-square test of independence to find out whether bachelor's degree has a relationship with racial identification, with racial identification being white, black, other for years 1993 and 2012. 
* * *

## Part III: Exploratory Data Analysis

Let's minimize our dataset to select variables that we require for our exploratory data analysis. 

```{r}
gss_eda <- gss %>% 
  select(caseid, year, age, sex, race, educ, paeduc, maeduc, speduc, coneduc, degree, region, partyid, relig, attend,conclerg, satfin,confed, conlegis, conjudge)
```

Now that we have all the variables necessary for our analysis let's start off with question 1.  

***Research Question 1:***

The first question deals with age, religion, attendance for religious services and confidence in religion. 

```{r}
gss_eda_1 <- gss_eda %>% 
  select(caseid, year, age, relig, attend, conclerg) %>% 
  filter(!is.na(age), !is.na(relig), !is.na(attend), !is.na(conclerg)) 

summary(gss_eda_1)
```

After running the initial summary of the data we find that apart from age and year, most of our variables of interest are factors such as attendance for religious services, religion and confidence in religious services.

####  (i) How has religious representation in the sample changed from 1973 to 2012. Create a line graph to show change overtime, and create a stacked barplot to show the change in each decade from 1973 to 2012. 

In order to see 10 year changes, let's first filter data to only observations in year 1972,1982,1992,2002,2012

```{r}
gss_eda_1_change <- gss_eda_1 %>% 
  filter(year %in% c(1973,1982,1993,2002,2012))
```

Now, let's summarise our data by creating a line graph showing the trend of religions identification in the United States and also make a bar graph to visualize 10 year changes in religious identification. 

```{r}
linegraphdata <- gss_eda_1 %>% 
  select(year, relig) %>% 
  group_by(year, relig) %>% 
  summarise(relig_count = n()) %>% 
      mutate(freq = relig_count / sum(relig_count))


ggplot(data = linegraphdata, mapping = aes(x = year, y = freq, color = relig)) +
  geom_line()+
    xlab("Year") +
  ylab("Religious Identification as a Proportion of Sample") +
   theme_economist()

religion_year_change <- gss_eda_1_change %>% 
  group_by(year, relig) %>% 
  summarise(count = n()) %>% 
    mutate(freq = count / sum(count))


ggplot(data = religion_year_change, aes(fill = relig, x = as.factor(year), y = freq)) +
  geom_bar(position = "stack",stat = "identity") + 
    xlab("Year") +
  ylab("Religious Identification as a Proportion of Sample") 

```

In Part 1 of Question 1, we find the proportion of people in the sample belonging to a particular religion, we create a line graph to represent a time series trend to show the the proportion of the sample belinging to a particular religion and how the representation changed from 1973 to 2012. We find that representation of people who were protestant decreased from around 63% to around 57%. Catholic representation remained the same while some religions that weren't a part of the sample in 1973 were included in the sample and increased overtime. These include Moslem, Orthodox Christian, Inter-nondenominational, Other and a few others. We filtered our data and reduced our sample to religions with at least 7 counts in order to make our visualization clearer. We also created a stacked bar chart to show how change in religious representation in year 1973,1982,2002 and 2012. We see that by 2002, most likely due to trends in immigration, we've seen an increase in other religions such as Orthodox Christianity, Moslem, Buddhism, Hinduism etc. Therefore, we find that religious representation of other religions has increased over the years most likely due to immigration. We believe reduction in Protestants in the sample could be due to two reasons, increase in other religions due to immigration and could also be due to more people affiliating themselves with no religion which can be seen in the chart as the people who are affiliated with no religion has increased. THerefore, over the years the United States has become more diverse in terms of religious representation and many people have also become unaffiliated with religion which may show the prevalance of Atheism. 

#### (ii) How has attendance in religious services changed between 1973 to 2013. Create a bar plot for attendance in religious services in 1973 and another in 2013. What differences do you see? 

```{r}
attendance <- gss_eda_1_change %>% 
  mutate(attend = as.character(attend),
         attend = if_else(attend == 'Lt Once A Year', '<1/YR', attend),
         attend = as.factor(attend)) %>% 
  mutate(attend = as.character(attend),
         attend = if_else(attend == 'More Thn Once Wk', '>1/WK', attend),
         attend = as.factor(attend)) %>% 
  mutate(attend = as.character(attend),
         attend = if_else(attend == 'Every Week', '1/WK', attend),
         attend = as.factor(attend)) %>% 
  mutate(attend = as.character(attend),
         attend = if_else(attend == 'Nrly Every Week', '<1/WK', attend),
         attend = as.factor(attend)) %>% 
  mutate(attend = as.character(attend),
         attend = if_else(attend == 'Once A Year', '1/YR', attend),
         attend = as.factor(attend)) %>% 
    mutate(attend = as.character(attend),
         attend = if_else(attend == 'Sevrl Times A Yr', '>1/YR', attend),
         attend = as.factor(attend)) %>% 
    mutate(attend = as.character(attend),
         attend = if_else(attend == '2-3X A Month', '2-3/MTH', attend),
         attend = as.factor(attend)) %>% 
    mutate(attend = as.character(attend),
         attend = if_else(attend == 'Once A Month', '1/MTH', attend),
         attend = as.factor(attend))

attendance_1973 <- attendance %>% 
  filter(year == 1973) %>% 
  select(attend) %>% 
  group_by(attend) %>% 
  summarise(count = n()) %>% 
  mutate(proportion = count/sum(count)*100)

summary(attendance_1973)

ggplot(data = attendance_1973, aes(x = attend, y = proportion, fill = attend)) +
  geom_bar(stat = "identity") +
      xlab("Religion") +
  ylab("Percentage of Responses") +
 scale_fill_manual(values = c("red", "grey", "seagreen3","purple","darkgreen","skyblue","darkblue","orange")) +
    geom_text(aes(label = round(proportion, digits = 2)),size = 3, position = position_stack(vjust = 0.5))


attendance_2012 <- attendance %>% 
  filter(year == 2012) %>% 
  select(attend) %>% 
  group_by(attend) %>% 
  summarise(count = n()) %>% 
  mutate(proportion = count/sum(count)*100)

summary(attendance_2012)

ggplot(data = attendance_2012, aes(x = attend, y = proportion, fill = attend)) +
  geom_bar(position = "stack",stat = "identity") +
        xlab("Religion") +
  ylab("Percentage of Responses")+
   scale_fill_manual(values = c("red", "grey", "seagreen3","purple","darkgreen","skyblue","darkblue","orange"))+
  geom_text(aes(label = round(proportion, digits = 2)),size = 3, position = position_stack(vjust = 0.5))

```

In this next step, we first reduce the length of characters in our categories for religious attendance, the aim here is to reduce characters in each category so the characters can fit well in our bar chart. We compare religious attendance in 1973 and 2012. We create charts for attendance in 1973,1982,2002, and 2012. We find that in 1973 that weekly religious attendance was the largest proportion with 23.9% respondents attending religious services weekly, with the second largest group being several times a year with around 17.5%, once a year bein around 15% and lowest proportion being once a month 7.3% for Once a month. In 2012, we find that weekly religious attendance has increased to 25%, several times a year has decreased to 12% while one a year has increased to 18% from 15%. More than once a week has increased from 8.69% to 10.2%. Therefore, we find that in some cases religious attendance has increased with weekly or more than once a week attendance percentages going up. On the other hand numbers for once a year has increased by 3%. Attendance 2-3 times a month has only increased by 2% while less than once a year has decrease by around 1-2%. Therefore we find that in some cases attendance has increased while in other cases it has decreased. 

#### (iii) How has confidence in organized religion changed, show decade long change with a stacked bar chart. 

```{r}
confidence_in_orgreligion <- gss_eda_1 %>% 
  select(year, conclerg) %>%
  group_by(year,conclerg) %>% 
  summarise(count = n()) %>% 
  mutate(Proportion_of_Confidence = count/sum(count)) %>% 
  filter(year %in% c(1973,1982,2002,2012))


ggplot(data = confidence_in_orgreligion, aes(fill = conclerg, x = as.factor(year), y = Proportion_of_Confidence)) +
  geom_bar(position = "stack",stat = "identity") +
        xlab("Year") +
  ylab("Proportion of Confidence in Organized Religion")+
  theme_economist() +
    geom_text(aes(label = round(Proportion_of_Confidence, digits = 2)),size = 3, position = position_stack(vjust = 0.5))

  
```

Now we analyze the figures for American confidence in organized religion. We find that from 1973 to 2002, the trend has changed. In 1973 almost 40% of Americans had a great deal of confidence in organized religion, while another 47 % had some confidence and 14% had hardly any confidence in organized religion. By 2002 these figures had changed with a large decrease in Americans who have a great deal in confidence in organized religion from 39% to 21%, while only some confidence increase by 13% to 60% and hardly any confidence increase 4% to 18%. By 2012, there was a slight increase in proportion of Americans who had a great of confidence in organized religion from 21% to 26%, and only some decreased by 3% and hardly any remained unchanged. Overall from 1973 to 2012 there has been an overall decrease in a great deal of confidence in organized religion while other categories have increased. 

***Research Question 2:***

Second question deals with education levels, bachelor's degree attainment, sex, and confidence in education. 


First let's select our data for this question and let's remove all NA values to simplify our analysis, 

```{r}
gss_eda_2 <- gss_eda %>% 
  select(caseid, year, age, sex, race, educ, coneduc, degree)

gss_eda_2 <- gss_eda_2 %>% 
filter(!is.na(age), !is.na(sex), !is.na(race), !is.na(educ),!is.na(coneduc), !is.na(degree))
```

#### (i) For simplicity, reclassify the degree variable into those Americans with bachelor's degree and those without a bachelor's degree. How has bachelor's degreee attainment changed in American from 1973 to 2012? Create a line graph and and a stacked bar chart to show this overtime change.  

Let's create a new variable which will be called bachelor_or_higher which will be a binary variable where Yes counts individuals with higher than a bachelor's degree and No accounts for individuals with less than a bachelor's degree and let's create a line plot showing the change in this trend from 1973 to 2012. 

```{r}
usbachelor <- gss_eda_2 %>% 
  select(caseid, year, degree,sex,race) %>% 
  mutate(bachelor_or_higher = case_when(degree=="Lt High School" ~ "No Bachelor's",
                                        degree=="High School" ~ "No Bachelor's",
                                        degree=="Junior College" ~ "No Bachelor's",
                                        degree=="Bachelor" ~ "Bachelor's or More",
                                        degree=="Graduate" ~ "Bachelor's or More"))

usbachelor_1 <- usbachelor %>% 
  select(year, bachelor_or_higher) %>% 
  group_by(year, bachelor_or_higher) %>% 
  summarise(count = n()) %>% 
  mutate(percentage_with_bachelor = count/sum(count))

ggplot(data = usbachelor_1, mapping = aes(x = year, y = percentage_with_bachelor, color = bachelor_or_higher)) +
  geom_line()+
    xlab("Year") +
  ylab("Percentage of People with Bachelor's Degree") +
   theme_economist()

usbachelor_10yr <- usbachelor %>% 
  select(year, bachelor_or_higher) %>% 
    filter(year %in% c(1973,1982,1993,2000,2012)) %>% 
  group_by(year, bachelor_or_higher) %>% 
  summarise(count = n()) %>% 
  mutate(percentage_with_bachelor = count/sum(count)*100)

usbachelor_10yr

ggplot(data = usbachelor_10yr, mapping = aes(x = as.factor(year) , y = percentage_with_bachelor, fill = bachelor_or_higher)) +
  geom_bar(stat = "identity")+
    xlab("Year") +
  ylab("Percentage with Bachelor's Degree") +
   theme_economist() +
 geom_text(aes(label = round(percentage_with_bachelor, digits = 2)),size = 3, position = position_stack(vjust = 0.5)) 
```

Both visualization show that percentage of Americans with bachelor's degree or higher has increased overtime. The line graph in 1973 starts with a large gap with a very small portion of Americans with bachelor's degrees or higher, overtime however, we see a convergence between the two lines No bachelor's and bachelor's or more. We find that overtime that number of people who have a bachelors degree of higher increases from somewhere around 10% to around 30% while the number of people with lower than bachelor's degree decreases from almost 90% in 1973 to 70% in 2012. The bar chart shows similar information but in 10 year intervels as proportion of people with bachelor's degree was relatively stagnant from 1973 to 1982 with around 10% people with bachelor's, then in 1993 we start to see the trend change as we see increase in bachelor's or more with almost 25% of the population having a degree. The figures remain around the same till 2000, and we see another 5% increase in 2012 to 30%. It is likely that as technological advancements took place and America shifted from manufacturing to services and technology, we see an increase in bachelor's degree. 

#### (ii) Have the proportion of Americans with bachelor degrees changed? Has the gendar ratio for attainment of a bachelor's degree changed from 1972 to 2012? Visualize this data. Have proportions of bachelor's degree attainment changed by race? Visualize this data. 

Now let's find if these numbers have changed based on gender in between 1973 and 2012. Let's first group and summarise our data based on year, bachelor's attainment and gender. 

```{r}
usbachelor_gender <- usbachelor %>% 
  select(year, bachelor_or_higher, sex) %>% 
     filter(year %in% c(1973,1982,1993,2000,2012)) %>% 
  group_by(year, bachelor_or_higher, sex) %>% 
  summarise(count = n()) %>% 
  ungroup(bachelor_or_higher) %>% 
  group_by(year, sex) %>% 
  mutate(percentage_with_bachelor = count/sum(count)*100)

usbachelor_gender

ggplot(usbachelor_gender, aes(x = sex, y = percentage_with_bachelor))+
  geom_bar(
    aes(fill = bachelor_or_higher), stat = "identity", color = "white",
    position = position_dodge(0.9)
    )+
      xlab("Year") +
  ylab("Bachelor's Attainment Based on Sex")+
  facet_wrap(~year) 
  
```

In this scenario we first summarize our data based on year, bachelor's degree and sex, and therefore we find the count of people with and without bachelor's degree or above and their sex. We then create a percentage for the number of women as a fraction of the total number of women with bachelor's degrees or higher and we compute the same for women. So the number of women with a without a bachelor's degree add up to 100. Now we create a bar plot for years 1973,1982,1993,2000 and 2012. We found previously that the percentage of individuals with bachelor's degree or higher was very low around 10% in 1973 and increased overtime to 30% in 2012. Here we find similar trends with the red bar showing bachelor's or more for both males and females and the blur bar showing lower than bachelor's for males and females. From 1973, we find that both men and women with bachelor's degree is very low, with women being lower than men, with men being around 15% and women being around 10%. Afterwards, we see an increase in 1982 for women to aroun 12% with men being around the same. We see a drastic increase in 1993, similar to our previous visualizations, We see a large increase for both men and women, with men with bachelor's degrees being around 28% of total men in the sample and women increasing from 12 to around 20%. The numbers remain around the same in 2002, however, by 2012 we see a slight increase for men of around a few percentage points, while women increased to almost around the same as men, probably a few percentage points lower. Therefore, from 1973 to 2012 we've seen an increase for both men and women, and ultimately gender parity when it comes to bachelor's degree or higher attainment, with men just having slightly higher attainment than women.

Now let's analyze bachelor's degree attainment based on race. 


```{r}
usbachelor_race <- usbachelor %>% 
  select(year, bachelor_or_higher, race) %>% 
  filter(year %in% c(1973,1982,1993,2000,2012)) %>% 
  group_by(year, bachelor_or_higher, race) %>% 
  summarise(count = n()) %>% 
  ungroup(bachelor_or_higher) %>% 
  group_by(year,race) %>% 
  mutate(percentage_with_bachelor = count/sum(count)*100) 


ggplot(usbachelor_race, aes(x = race, y = percentage_with_bachelor))+
  geom_bar(
    aes(fill = bachelor_or_higher), stat = "identity", color = "white",
    position = position_dodge(0.9)
    )+
      xlab("Year") +
  ylab("Bachelor's Attainment Based on Race")+
  facet_wrap(~year) 
```

Now, let's analyze how the bachelor's degree attainment has changed based on the race between 1973 and 2012. In this scenario we first summarize our data based on year, bachelor's degree or higher and race, and therefore we find the count of people with and without bachelor's degree or above and their race. We then create a percentage for the number of people with a bachelor's degree or without a bachelor'r degree based on their race. For example total percentage of white people with bachelor's degree or higher and total number of white people without a bachelor's degree add up to 100. Now we create a bar plot for years 1973,1982,1993,2000 and 2012. We are going to ignores figures for "Other" race category in 1973. We found that there were less than 10 respondents for other category which can cause sampling errors so it important to ignore these numbers from our sample. In 1973 we found that around 14% of white respondents had a bachelor's or higher and less than 9% of black respondents had a bachelor's or higher. In 1982 we find that number had only slightly increase for white respondents but increased for black and other respondents to almost 12%. In 1993 we see numbers increase for White and Other but actually reduce slightly for black respondents. We see increase for white respondents from 12% to 25% and for other from 11% to 31%.In 2000 we see little changes for any category and actually a decrease for the other category from 31% to 25%. By 2012, we see the the number of college degree or higher holding black respondents increase drastically from 12% to 20%. We see a 5% increase for White Americans from 26% to around 31% and a 4% increase for other from 25% to 29%.

#### (iii) Is there a relationship between confidence in education and education levels. Explore both variables, visualize the relationship between these variables and also create a correlation matrix for years 1973, 2000 and 2012.  

Lastly, we find out if there a relationship between confidence in education and education levels. We do this by visualizing this data and also calculating the correlation between these two variables. 

```{r}
useducation <- gss_eda_2 %>% 
  select(year, educ, coneduc)

useducation_1 <- useducation %>% 
  mutate(coneduc = as.character(coneduc),
         coneduc = if_else(coneduc == 'A Great Deal', '3', coneduc),
         coneduc = as.factor(coneduc)) %>% 
  mutate(coneduc = as.character(coneduc),
         coneduc = if_else(coneduc == 'Only Some', '2', coneduc),
         coneduc = as.factor(coneduc)) %>% 
  mutate(coneduc = as.character(coneduc),
         coneduc = if_else(coneduc == 'Hardly Any', '1', coneduc),
         coneduc = as.factor(coneduc))

useducation_1 <- useducation_1 %>% 
  select(year, educ, coneduc) %>% 
  group_by(year) %>% 
  summarise(mean_education = mean(educ), mean_coneduc = mean(as.numeric(coneduc)))


ggplot(data = useducation_1, mapping = aes(x = year, y = mean_education, color = mean_education)) +
  geom_line()+
    xlab("Year") +
  ylab("Level of Education")

ggplot(data = useducation_1, mapping = aes(x = year, y = mean_coneduc, color = mean_coneduc)) +
  geom_line()+
    xlab("Year") +
  ylab("Level of Confidence in Education")


```

Firsly, we turn confidence in education into a numeric variable, coding 1 for individuals who choose "Hardly Any" confidence in education, 2 for "Only Some" Confidence in Education and 3 for "A Great Deal" of Confidence in Education. Afterwards we find mean education levels and mean confidence in education from 1973 to 2012. We find that mean education level has increased overtime from 11.8 to 13.6. From 1973, we see volatility with increases and decreases, but overall increase till 1984, then we see a smooth increase from 1984 to 1996, after which we see volatility between 1996 and 2004 but we see an overall increases, we see a sharp decline from 2004 to 2006, after which we an overall increase till 2012. We see that confidence in education has decreased overtime from 2.3 in 1973 to around 2.09 in 2012. We see a lot of volatility with sharp declines in the 1970s and 1980s after we see some increases in the early 1990s till 2006, after which we see another sharp decline till 2012. We do an improvement from the really low confidence in education in 1993 of 2.05 to to 2.09 in 2012. By 2012, most Americans have only a little greater than "Only Some" Confidence in Education which was coded as 2 when we converted our factors to numeric variables. Therefore, we see that overall education levels have increased from 11.8 to 13.6, we have seen confidence in education has decreased from 2.3 to 2.09, therefore, the two variables seem to be negatively correlated overtime. 

```{r}
useducation_2 <- useducation %>%
  select(year, coneduc) %>% 
  group_by(year, coneduc) %>% 
  summarise(count = n()) %>% 
  mutate(percent_coneduc = count/sum(count)) %>% 
  filter(year %in% c(1973,1982,1993,2000,2012))

ggplot(data = useducation_2, mapping = aes(x = factor(year) , y = percent_coneduc, fill = coneduc)) +
  geom_bar(stat = "identity")+
    xlab("Year") +
  ylab("Percentage of Confidence in Education") +
 geom_text(aes(label = round(percent_coneduc, digits = 2)),size = 3, position = position_stack(vjust = 0.5))
```

After finding the averages for education and confidence in education, we want to analyze the data based on our categories, finding out how Americans responded in all 3 categories of confidence in education. We also want to see if there is a correlation between education level and confidence in education across all 3 categories, meaning if response for confidence in education differs by education level. Therefore, first we create a bar plot comparing the percentage of respondents of the total respondents that answered in each of the three categories for confidence in education, "A Great Deal", "Only Some" and "Hardly Any" confidence in education. In 1973 we find that only 8% of Americans had "Hardly Any" confidence in education, while 37% had "A Great Deal" of confidence in education and more than half had "Only Some" confidence in education. We find that over the next few decades confidence in education decreases, as individuals who had "A Great Deal" of cofidence in education declines to only 23% in 1993, and indivuduals with "Hardly Any" confidence in education increases from 8% in 1973 to 18% in 1993 with almost 60% Americans having "Only Some" confidence in education. By 2012, these figures have changed only slightly as individuals with "A Great Deal" of confidence in education have increased 3 percentage points to 26% and individuals with "Hardly Any" confidence in education have decreased 1% to 17% and "Only Some" has decreased a few percentage points to around 58%, remaining the majority of respondents choice between 1973 and 2012. Therefore, overtime we have seen confidence in education decrease overall, as we saw similarly with our mean in the line graph. We did, however, see a slight trend upward between 1993 to 2012 but overall decrease from 1973. 

```{r}
useducation_3 <- useducation %>% 
  mutate(coneduc = as.character(coneduc),
         coneduc = if_else(coneduc == 'A Great Deal', '3', coneduc),
         coneduc = as.factor(coneduc)) %>% 
  mutate(coneduc = as.character(coneduc),
         coneduc = if_else(coneduc == 'Only Some', '2', coneduc),
         coneduc = as.factor(coneduc)) %>% 
  mutate(coneduc = as.character(coneduc),
         coneduc = if_else(coneduc == 'Hardly Any', '1', coneduc),
         coneduc = as.factor(coneduc)) 


useducation_3_1 <- useducation_3  %>% 
  select(year, educ, coneduc) %>% 
  filter(year %in% c(1973,1982,1993,2000,2012)) %>% 
  group_by(year,coneduc) %>% 
  summarise(mean_education = mean(educ))
  

ggplot(data = useducation_3_1, mapping = aes(x= coneduc, y = mean_education, fill = coneduc)) +
  geom_bar(stat = "identity")+
  facet_wrap(~year)+
    xlab("Year") +
  ylab("Level of Confidence in Education by Mean Education Levels") +
 geom_text(aes(label = round(mean_education, digits = 2)),size = 3, position = position_stack(vjust = 0.5))
  
  
useducation_3_4 <- useducation_3 %>% 
  select(year, educ, coneduc) %>% 
  filter(year == 1973)

useducation_3_4$coneduc <- as.integer(useducation_3_4$coneduc)

useducation_cor_1973 <- cor.test(useducation_3_4$educ, useducation_3_4$coneduc, 
                    method = "pearson")

useducation_cor_1973 


useducation_3_6 <- useducation_3 %>% 
  select(year, educ, coneduc) %>% 
  filter(year == 2000)

useducation_3_6$coneduc <- as.integer(useducation_3_6$coneduc)

useducation_cor_2000 <- cor.test(useducation_3_6$educ, useducation_3_6$coneduc, 
                    method = "pearson")

useducation_cor_2000

useducation_3_5 <- useducation_3 %>% 
  select(year, educ, coneduc) %>% 
  filter(year == 2012) 


useducation_3_5$coneduc <- as.integer(useducation_3_5$coneduc)

useducation_cor_2012 <- cor.test(useducation_3_5$educ, useducation_3_5$coneduc, 
                    method = "pearson")

useducation_cor_2012
```
Now we analyze whether there is a relationship between education levels and confidence in education. We create a barplot for the mean education levels of the respondents for each category, "A Great Deal" coded for 3, "Only Some" coded for 2, and "Hardly Any" coded for 1. We find the mean education levels for each of the respondents in these categories for years 1973, 1982, 1993, 2000 and 2012. We find that education levels for each category have increased overtime from 1973 to 2012. We find that mean education was between 11 to 12.2 for each category in 1993, by 2012 the mean education for each category was between 13 to almost 14. We find that individuals that had "A Great Deal" of confidence in education had the lowest level of education across 3 groups for each year, meaning individuals with "A Great Deal" of confidence in education are less education than other categories. However, by 2000 we find that mean education for all categories has almost converged with individuals who responded in each category having roughly around same education levels between 13.2 and 13.4, we see slight divergence in 2012, with individuals with the most mean education level of 13.85 having "Hardly Any" confidence in education. Overall we find that in 1973 that those with the lowest education level of 11.27 across these categories had "A Great Deal" of confidence in education, while those with "Hardly Any' Confidence in Education had mean education level of 11.39. By 2012, as education levels increased, again  those individuals who had "A Great Deal" of confidence in education had a mean education level of 13, while those who had a great deal of confidence in education. 

Finally we calculate the pearson correlation coefficient for years 1973, 2000 and 2012 to show the relationship between education levels and confidence in education. We find that in 1973 the coefficient was -0.0784 which means that there is a negative relationship between education level and confidence in education. Meaning a higher education level would mean a lower confidence in education and this figure is significant with a p-value less than 0,05. For year 2000 we find the correlation coefficient to be -0.0166 which is less correlated than 1973, however, our p-value is 0.4757 which is very higher, much higher than the 0.05 so this estimate may not be reliable and we can see that the confidence intervel includes 0.  However, in 2012 we see a large coefficient of -0.096 which shows an even larger negative relationship between education levels and confidence in education than 1973. Therefore, as education levels increase, confidence in education decreases. This estimate is significant with a p-value much lower than 0.05. Therefore we seem to see a negative correlation between education levels and confidence in education, showing that more educated people have less confidence in education. Our coefficient is below 0.1 for all years so this correlation is not very strong but is significant at least for years 1973 and 2012.  

* * *

## Part IV: Statistical Inference

In order to showcase skills in both hypothesis testing for mean, hypothesis test for comparing two dependent mean, comparing more than two means, hypthesis test for proportions, comparing two proportions,and evaluate independence of two categorical variables with a chi-squared test of independence. we will do one question for proportion testing, one question for mean hypothesis testing and one question for group testing. 

***Research Question 1: ***

In this first research question, we will do analysis on education levels in the US. We will compare education levels in the US to other countries, find out whether education levels in the US differ by gender, and lastly find out if education levels in the US differ by race. 

### Part 1: Hypothesis Test of Means 

(i) We find that the mean education in Australia in 1975 was 10.04 and the mean education for Canada in 1975 was 9.18.Conduct a hypothesis test for means, finding out if the mean education in the United States in 1973 is greater than the mean education in Australia, conduct simialr hypothesis test for US and Canada. Lastly find the confidence intervel for mean education level in the United States. 


First let's check our conditions. 

```{r}
gss_inference_1 <- gss_eda %>% 
  select(year, educ) %>% 
  filter(!is.na(educ)) %>% 
  filter(year %in% c(1973))

mean(gss_inference_1$educ)

hist(gss_inference_1$educ)

nrow(gss_inference_1)
```

We find that our point estimate is 11.59. For our conditions, we find that n > 30 and our histogram shows that our distribution is nearly normal. We know from our data portion that is this the general social survey and random sampling is used in this survey. Therefore, our conditions are met. Now let's conduct our hypothesis test. 

```{r}
gss_inference_1 <- gss_eda %>% 
  select(year, educ) %>% 
  filter(!is.na(educ)) %>% 
  filter(year %in% c(1973))

mean(gss_inference_1$educ)

# Test for Australia
inference(y = educ, data = gss_inference_1, statistic = "mean", type = "ht", null = 10.04, 
          alternative = "greater", method = "theoretical")

#Test for Canada
inference(y = educ, data = gss_inference_1, statistic = "mean", type = "ht", null = 9.18, 
          alternative = "greater", method = "theoretical")
inference(y = educ, data = gss_inference_1, statistic = "mean", type = "ci", method = "theoretical")
```

In both cases we find that the p-value is lower than 0,05 and therefore, we reject the null hypothesis that the mean education level of US is equal to that of Canada or Australia. We also find the confidence intervel for mean education in the US. We are 95% confident that the true mean education in the US is between 11.4284 and 11.7657. 

### Part 2: Confidence Intervals for Two Means 
(ii)Is there convincing evidence that the United States has seen a change in its education levels from 1973 to 2012. 

```{r}
gss_inference_1_2 <- gss_eda %>% 
  select(year, educ) %>% 
  filter(!is.na(educ)) %>% 
  filter(year %in% c(2012))

hist(gss_inference_1$educ)
hist(gss_inference_1_2$educ)
inference(y = educ ,data =  gss_inference_1, statistic = "mean", type = "ci", method = "theoretical", conf = 0.95)
inference(y = educ,data =  gss_inference_1_2, statistic = "mean", type = "ci", method = "theoretical", conf = 0.95)
```

Since n>30 and distribution is nearly normal for 1973. Distribution is left skewed for 2012 but since we have a large sample, we believe our distibution is nearly normal. We also know that sample is less than 10% of the population. Since both our confidence intervels don't overlap, we find that the United States has seen a significant change in its education levels. The 95% confidence intervals for education levels in 1973 is (11.4284 , 11.7657) and the 95% confidence intervals for education levels in 2012 is (13.3898 , 13.666). 

### Part 3: Hypothesis Test for Two Independent Means 
(iii) We would like to esimtate the difference between two independent means. Therefore, conduct a hypothesis test evaluating whether the education levels of male is different than the education level of female. Find inference for both 1973 and 2012. 


```{r}
gss_inference_1_3_1 <- gss_eda %>% 
  select(year, educ, sex) %>% 
  filter(!is.na(educ), !is.na(sex)) %>% 
  filter(year %in% c(1973))

gss_inference_1_3_2 <- gss_eda %>% 
  select(year, educ, sex) %>% 
  filter(!is.na(educ), !is.na(sex)) %>% 
  filter(year %in% c(2012))

inference(y = educ, x = sex, data = gss_inference_1_3_1, statistic = "mean", type = "ht", method = "theoretical", null = 0, alternative = "twosided")
inference(y = educ, x = sex, data = gss_inference_1_3_2, statistic = "mean", type = "ht", method = "theoretical", null = 0, alternative = "twosided")
```

After conducting hypothesis test, we find that the p-value in both cases is above 0.05. Therefore, we fail to reject the null hypothesis that there is no significant difference in male and female education level in both 1973 (p = 0.2059) and 2012 (p = 0.7766) 

### Part 4: Analaysis of Variance: Education and Race

(iv) Let's conduct one more test comparing more than two means, in this case an ANOVA. Let's find out whether education levels are different based on race. Let's conduct this for both year 1973 and year 2012. 

Our null hypothesis will be that the mean education level is same across all 3 race categories while our alternative hypothesis is that atleast one pair of means is different from each other. We will calculate the F-Statistic in this case. In order to reject the null hypothesis we need a small p-value (less than 0.05).

```{r}
gss_inference_1_4_1 <- gss_eda %>% 
  select(year, educ, race) %>% 
  filter(!is.na(educ), !is.na(race)) %>% 
  filter(year %in% c(1973))

gss_inference_1_4_2 <- gss_eda %>% 
  select(year, educ, race) %>% 
  filter(!is.na(educ), !is.na(race)) %>% 
  filter(year %in% c(2012))

summary(gss_inference_1_4_1)
summary(gss_inference_1_4_2)

#filter our other in 1973
gss_inference_1_4_1 <- gss_inference_1_4_1 %>% 
  filter(race != "Other") 

gss_inference_1_4_1$race <- factor(gss_inference_1_4_1$race)
 
summary(gss_inference_1_4_1)

ggplot(data = gss_inference_1_4_1, mapping = aes(y = educ, x = race, fill = race))+
  geom_boxplot()

ggplot(data = gss_inference_1_4_2, mapping = aes(y = educ, x = race, fill = race))+
  geom_boxplot()

ggplot(data = gss_inference_1_4_1, mapping = aes(y = educ, fill = race))+
  geom_histogram()+
  facet_wrap(~race)

ggplot(data = gss_inference_1_4_2, mapping = aes(y = educ, fill = race))+
  geom_histogram()+
  facet_wrap(~race)

inference(y = educ, x = race, data = gss_inference_1_4_1, statistic = "mean", type = "ht", method = "theoretical", null = 0, alternative = "twosided")

ANOVA_results_2012 <- aov(educ ~ race, data = gss_inference_1_4_2)

summary(ANOVA_results_2012)
```

In  order to meet the conditions of ANOVA. We must assume within group and between group independence. Between group independence is met since observations are independent as we took a random sample and we know that sample of less than 10% of the population and we have sampled without replacement and observations are non-paired. We see in our histogram that distributions are left skewed, but since we have a large number of observations we can assume approximate normality. Now we need to look at the within group variability. We create boxplots and find that variability is very high for "other" race in 1973 due to the number of observations for other group being very low. Therefore, we should remove observations in other group in 1973 since within group variability is too high and only do a standard hypothesis test for inference for two means and conduct an ANOVA for 2012 data. 

For 1973, we conduct a hypothesis test and find a highly significant p-value of less than 0,01. Therefore, we reject the null hypothesis and have sufficient evidence that there is a significant difference between education levels between white and black race categories. For 2012, we find the F-statistic for year 2012. The first row which shows between group variability since race is the category across which we are comparing our means. The second row is within group variability, which is also known as the error row. Our first row of explained sum of squares which is the variability in the response variable explained by variability in the analysis i.e. between group variability. The second row of residuals is the Sum of Squares Errors (SSE) which measures the within group variability i.e. the variability unexplained by the group variability or due to other reasons. We divide our Sum of Squares with our degrees of freedom and we get our mean squares, and we divide our Mean Squared Group by Mean Squared Errors to get our F-Statistics which is 10.21 for 2012. We use our F-Statistic to calculate our p value which in this case is defined by the probability of an observed or more extreme outcome given the null hypothesis is true. We find our p-value is really low and therefore we reject the null hypothesis and that we find convincing evidence that at least one pair of means of the race group are different from each other. Now we conduct a pairwise comparison. 

Our ANOVA results show that at least one pairs of means is different but we don't know which one. Therefore, we conduct a pairwise comparison to find out. 

```{r}
TukeyHSD(ANOVA_results_2012)

pairwise.t.test(gss_inference_1_4_2$educ, gss_inference_1_4_2$race, p.adjust.method = "BH")
pairwise.t.test(gss_inference_1_4_2$educ, gss_inference_1_4_2$race, p.adjust.method = "bonf")
```

We conduct the Tukey pairwise comparison and find that all pairwise differences except Other-Black are significant with a p-value lower than 0.05. We also do a pairwise t-test with the Benjamini-Hochberg method and the more strict Bonferroni method and find that all groups are significant except for Other-Black which seems to be the only insignificant pairwise comparison. Therefore, we find that education levels are much higher for people who identified as White compared to other race categories. Therefore, there seems to be a correlation between race and education levels. 

***Research Question 2:***

In Question 2 we will conduct hypothesis test for proportions, estimate the difference between two proportions, conduct a hypothesis test comparing two sample proportions and and test the independence of two categorical variables, with at least one of the variables having more than one level. 

### Part 1: Hypothesis Test for Proportion and Difference Between Two Proportions

(i) According to sources, 22.44% of Canadians and 30% Germans have bachelor's degrees. Conduct a hypothesis test for proportions and find if the proportion of Americans with bachelor's degrees is greater than Canadians and Germans in year 2012. 

For Part 1 (i) let's state both or null and alternative hypothesis. 

(i) Null hypothesis is that American bachelor's degree attainment is less than Canada
(ii) Alternative hypothesis states that American bachelor's degree is higher than Canada


```{r}
gss_inference_2_1_1 <- gss_eda %>% 
  select(year, degree)

gss_inference_2_1_1 <- gss_inference_2_1_1 %>% 
  select(year, degree) %>% 
  mutate(bachelor_or_higher = case_when(degree=="Lt High School" ~ "No Bachelor's",
                                        degree=="High School" ~ "No Bachelor's",
                                        degree=="Junior College" ~ "No Bachelor's",
                                        degree=="Bachelor" ~ "Bachelor's or More",
                                        degree=="Graduate" ~ "Bachelor's or More"))

gss_inference_2_1_1 <- gss_inference_2_1_1 %>% #remove NAs and filter out all years except 2012
  filter(!is.na(degree),!is.na(bachelor_or_higher)) %>% 
  filter(year == "2012") 

gss_inference_2_1_1$bachelor_or_higher <- as.factor(gss_inference_2_1_1$bachelor_or_higher)
summary(gss_inference_2_1_1) 

gss_inference_2_1_1 %>% 
  select(bachelor_or_higher) %>% 
  group_by(bachelor_or_higher) %>% 
  summarise(count = n()) %>% 
  mutate(proportion = count/sum(count)*100)

```

Now that we've arranged our data in our preferred manner let's see if our data meets the condition for a hypothesis test of proportions. We know this is survey data and some sort of random sampling method was applied so observations are independent. We know that the proportions for bachelor's or more and no bachelor's is 28.433% and 71.56% respective. Let's see if we meet the success/failure condition. The condition is np>10 and n(1-p)>10 where p is success which in this case is bachelor's degree attainment. 

```{r}
559+1407

0.28*1966>10
0.7156*1966>10
```

Since both our results are greater than 10, we have met the success failure condition. We will now conduct a hypothesis test. 

```{r}
#Canada
inference(y = bachelor_or_higher, data = gss_inference_2_1_1, statistic = "proportion", type = "ht", alternative = "greater", null = 0.224,  method = "theoretical", success = "Bachelor's or More")

#Germany 
inference(y = bachelor_or_higher, data = gss_inference_2_1_1, statistic = "proportion", type = "ht", alternative = "greater", null = 0.30,  method = "theoretical", success = "Bachelor's or More")
```

Therefore, we see our results for Canada and we reject the null hypothesis and find there is convincing evidence that American bachelor's degree attainment is greater than Canada. In the case of Germany, we fail to reject the null hypothesis and find that there is convincing evidence that American bachelor's degree attainment is not greater than Germany. 

### Part II 

(ii) Is there convincing evidence that United States has seen a change in its bachelor's degree attainment levels between 2002 and 2012?

For (ii) we want to find out if America has seen a change in its bachelor's degree attainment levels from 2002 and 2012. Since we are dealing with similar data we find that we have met all our conditions. So we arrange our data and calculate our confidence intervals. 

```{r}
gss_inference_2_1_2 <- gss_eda %>% 
  select(year, degree)

gss_inference_2_1_2 <- gss_inference_2_1_2 %>% 
  select(year, degree) %>% 
  mutate(bachelor_or_higher = case_when(degree=="Lt High School" ~ "No Bachelor's",
                                        degree=="High School" ~ "No Bachelor's",
                                        degree=="Junior College" ~ "No Bachelor's",
                                        degree=="Bachelor" ~ "Bachelor's or More",
                                        degree=="Graduate" ~ "Bachelor's or More"))

gss_inference_2_1_2_1 <- gss_inference_2_1_2 %>% 
  filter(!is.na(degree),!is.na(bachelor_or_higher)) %>% 
  filter(year == "2002") %>% 
  select(year, bachelor_or_higher)

gss_inference_2_1_2_2 <- gss_inference_2_1_2 %>% 
  filter(!is.na(degree),!is.na(bachelor_or_higher)) %>% 
  filter(year == "2012") %>% 
  select(year, bachelor_or_higher)

inference(y = bachelor_or_higher, data = gss_inference_2_1_2_1, statistic = "proportion", type = "ci",  method = "theoretical", success = "Bachelor's or More")

inference(y = bachelor_or_higher, data = gss_inference_2_1_2_2, statistic = "proportion", type = "ci",  method = "theoretical", success = "Bachelor's or More")

gss_inference_2_1_2_3 <- gss_inference_2_1_2 %>% 
  filter(!is.na(degree),!is.na(bachelor_or_higher)) %>% 
    filter(year %in% c(2002,2012)) %>% 
  select(year, bachelor_or_higher)

inference(y = bachelor_or_higher, x = year, data = gss_inference_2_1_2_3, null = 0, statistic = "proportion", type = "ht",  method = "theoretical", success = "Bachelor's or More", alternative = "twosided")
```

We calculate confidence intervals for years 2002 and 2012 and find that these confidence intervels don't overlap. Since they don't overlap, we have convincing evidence that the United States has indeed seen a significant change in bachelor's degree attainment from 2002 to 2012. Confidence interval for 2002 is CI: (0.229 , 0.2612) and for 2012 is CI: (0.2644 , 0.3043).

### Part III 

(iii) Conduct a hyppthesis test of two sample proportions and find if there is a relationship between bachelor's degree attainment and party ID (for simplicity we will assume that there are only two parties: Democratic and Republican Party and delete all other observations). We will conduct a confidence intervel to estimate whether there is a difference between democrats with a bachelor's degree and republican with a bachelor's degree for years 1993 and 2012 

```{r}
gss_inference_2_1_3_1 <- gss_eda %>% 
  select(year, degree, partyid) %>% 
  filter(!is.na(year),!is.na(degree),!is.na(partyid))
summary(gss_inference_2_1_3_1$partyid)

gss_inference_2_1_3_1 <- gss_inference_2_1_3_1 %>% 
mutate(bachelor_or_higher = case_when(degree=="Lt High School" ~ "No Bachelor's",
                                        degree=="High School" ~ "No Bachelor's",
                                        degree=="Junior College" ~ "No Bachelor's",
                                        degree=="Bachelor" ~ "Bachelor's or More",
                                        degree=="Graduate" ~ "Bachelor's or More"))
gss_inference_2_1_3_1 <- gss_inference_2_1_3_1 %>% 
  mutate(partyid_2 = case_when(partyid=="Strong Democrat" ~ "Democrat",
                                        partyid=="Not Str Democrat" ~ "Democrat",
                                        partyid=="Ind,Near Dem" ~ "Democrat",
                                        partyid=="Ind,Near Rep" ~ "Republican",
                                        partyid=="Not Str Republican" ~ "Republican",
                                        partyid=="Strong Republican" ~ "Republican",
                                        partyid=="Independent"~ "NA", 
                                        partyid=="NA" ~ "NA",
                                        partyid=="Other Party" ~ "NA")) %>% 
  mutate(partyid_2 = as.factor(partyid_2))

gss_inference_2_1_3_1 <- gss_inference_2_1_3_1 %>% 
  filter(partyid_2!="NA") 

summary(gss_inference_2_1_3_1$partyid_2)

gss_inference_2_1_3_1$partyid_2 <- droplevels(gss_inference_2_1_3_1$partyid_2)

summary(gss_inference_2_1_3_1$partyid_2)

gss_inference_2_1_3_2 <- gss_inference_2_1_3_1 %>% 
  filter(year == "1993")

gss_inference_2_1_3_3 <- gss_inference_2_1_3_1 %>% 
  filter(year == "2012")
```

For simplicity we'll group all categories of democrats into one and all categories of republicans into one and we'll delete all independents. Now that we've rearranged our data, let's state our hypothesis. Our null hypothesis states that there is no relationship between partyid and bachelor's degree meaning that proportion of democrats with a bachelor's degree is equal to proportion of republicans with bachelor's degree. Our alternative hypothesis states that there is a difference between proportion of republican and proportion of democrats with bachelor's degrees. Now let's check our conditions. We know that our survey data is collected using some sort of random sampling, we know that our sample without replacement is less than 10% of the population. 

```{r}
summary(gss_inference_2_1_3_2)

gss_inference_2_1_3_2 %>% 
  group_by(year, bachelor_or_higher, partyid_2) %>% 
  summarise(count = n()) %>% 
  ungroup() %>% 
  group_by(partyid_2) %>% 
  mutate(proportion = count/sum(count)*100)


#Bachelor's Degree Success Failure Condition
#Democrats
155+570
0.21*725>10
570*0.7862>10
#Republicans 
443+185
0.2945*628>10
0.7054140*628>10



gss_inference_2_1_3_3 %>% 
  group_by(year, bachelor_or_higher, partyid_2) %>% 
  summarise(count = n()) %>% 
    ungroup() %>% 
  group_by(partyid_2) %>% 
  mutate(proportion = count/sum(count)*100)

#Bachelor's Degree Suceess Failure Condition
#Democrats 
285+647
0.3057*932>10
0.6942*932>10
#Republican
189+409
0.31605*598>10
0.6839*598>10
```



```{r}
inference(y = bachelor_or_higher, x = partyid_2, data = gss_inference_2_1_3_2, null = 0, statistic = "proportion", type = "ci",  method = "theoretical", success = "Bachelor's or More", alternative = "twosided")

inference(y = bachelor_or_higher, x = partyid_2, data = gss_inference_2_1_3_2, null = 0, statistic = "proportion", type = "ht",  method = "theoretical", success = "Bachelor's or More", alternative = "twosided")

inference(y = bachelor_or_higher, x = partyid_2, data = gss_inference_2_1_3_3, null = 0, statistic = "proportion", type = "ci",  method = "theoretical", success = "Bachelor's or More", alternative = "twosided")

inference(y = bachelor_or_higher, x = partyid_2, data = gss_inference_2_1_3_3, null = 0, statistic = "proportion", type = "ht",  method = "theoretical", success = "Bachelor's or More", alternative = "twosided")
```

After estimating difference between two proportions we find that for year 1993, the confidence intervals show that there is a significant difference between bachelor's degree attainment for democrats and republicans. Our confidence intervals runs negative and we have a 95% confidence interval of 95% CI (Democrat - Republican): (-0.1273 , -0.0343). Since our confidence intervals does not include 0, we find that this negative difference is significant. We can verify that by conducting a hypothesis test which is highly significant with a p-value much lower than 0.05. Therefore, there is a significant difference between bachelor's degree attainment of republicans and democrats, as attainment is higher for republicans, we reject the null hypothesis. 

After estimating difference between two proportions we find that for year 2012, the confidence intervals show that there is not a significant difference between bachelor's degree attainment for democrats and republicans. We have a 95% confidence interval of 95% CI (Democrat - Republican): (-0.0578 , 0.0373). Since our confidence intervals does include 0, we find that this difference is insignificant. We can verify that by conducting a hypothesis test which is highly insignificant with a p-value greater than 0.05. Therefore, we fail to reject the null hypothesis, there is not a significant difference in bachelor's degree attainment for democrats and republicans.

### Part 2: Chi-Square Test of Independence. 

Conduct a chi-square test of independence to find out whether bachelor's degree has a relationship with racial identification, with racial identification being white, black, other for years 1993 and 2012. 

Our null hypothesis states that there is no relationship between bachelor's degree attainment and race, our alternative hypothesis states that there is a relationship between bachelor's degree attainment and race. 

Let's first check our conditions. We know this survey uses random sampling method and samples less than 10% of the population without replacement.

```{r}
gss_inference_2_2_1_1 <- gss_eda %>% 
  select(year, degree, race) %>% 
  filter(!is.na(degree),!is.na(race)) %>% 
   mutate(bachelor_or_higher = case_when(degree=="Lt High School" ~ "No Bachelor's",
                                        degree=="High School" ~ "No Bachelor's",
                                        degree=="Junior College" ~ "No Bachelor's",
                                        degree=="Bachelor" ~ "Bachelor's or More",
                                        degree=="Graduate" ~ "Bachelor's or More")) %>%   mutate(bachelor_or_higher = as.factor(bachelor_or_higher))

summary(gss_inference_2_2_1_1$bachelor_or_higher)

gss_inference_2_2_1_2 <- gss_inference_2_2_1_1 %>% 
  filter(year == "2002")

gss_inference_2_2_1_3 <- gss_inference_2_2_1_1 %>% 
  filter(year == "2012")

gss_inference_2_2_1_2 %>% 
  select(bachelor_or_higher, race) %>% 
  group_by(bachelor_or_higher, race) %>% 
  summarise(count = n()) %>% 
  ungroup() %>%
  group_by(race) %>% 
  mutate(prop = count/sum(count)*100)

gss_inference_2_2_1_3 %>% 
  select(bachelor_or_higher, race) %>% 
  group_by(bachelor_or_higher, race) %>% 
  summarise(count = n()) %>% 
  ungroup() %>%
  group_by(race) %>% 
  mutate(prop = count/sum(count)*100)
```

Another condition is that each expected cell have at least 5 cases, which is easily met. We also know that each case only contributes to one cell in the table. 

What is the bachelor's degree attainment rate in the sample. 

```{r}
gss_inference_2_2_1_2 %>% 
  select(bachelor_or_higher) %>% 
  group_by(bachelor_or_higher) %>% 
  summarise(count = n()) %>% 
  mutate(prop = count/sum(count)*100)

gss_inference_2_2_1_3 %>% 
  select(bachelor_or_higher) %>% 
  group_by(bachelor_or_higher) %>% 
  summarise(count = n()) %>% 
  mutate(prop = count/sum(count)*100)
```

Attainment rate is 24.5% in 2002 and 28.4% in 2012. If the null hypothesis is true the bachelor's degree attainment rate should be the same for all races: white,black and other. Let's find our hypothesized counts if the null hypothesis is true. Let's do these calculations just for 2002 as an example and assume a similar procedure for 

```{r}
#Total * average attainment rate

#White
446+1025
0.25*1471 #should be attainment for white if null is true


#Black
60+240
300*0.25 #Attainment for blacks if null true 

#Other
53+142
0.25*195 #Attainment for others if null true 
```

Now let's find our solution by using the chi-square test of independence. 

```{r}
tbl = table(gss_inference_2_2_1_2$bachelor_or_higher, gss_inference_2_2_1_2$race)
tbl
chisq_2002 <- chisq.test(tbl)
chisq_2002
chisq_2002$residuals
tbl_1 = table(gss_inference_2_2_1_3$bachelor_or_higher, gss_inference_2_2_1_3$race)
tbl_1
chisq_2012 <- chisq.test(tbl_1)
chisq_2012
chisq_2012$residuals
```

In both cases. we have a high chi-square statistic and low p-value for years 2002 and 2012. Therefore, we reject the null hypothesis, there is convincing evidence that bachelor's degree attainment is not independent of race in both years. However, we do find that the residual differences have decreased overtime from 2002 to 2012. And chi-sq value is much higher for 2002 than 2012 and p-value is lower for 2002 than 2012. 

```{r}
library(corrplot)
corrplot(chisq_2002$residuals, is.cor = FALSE)
corrplot(chisq_2012$residuals, is.cor = FALSE)

gss_inference_pairwise <- gss_inference_2_2_1_2 %>% 
  select(bachelor_or_higher, race)
```

We visualize our pearson residuals and we find that these residuals are highest for black and white races while "other" race seems to have the lowest residuals and therefore is closest to the average degree attainment. Residuals are highest for individuals who identify as black, then white, with lowest residuals being for individuals who identify as other. The patterns are similar for both 2002 and 2012 with residuals being higher 2002 than 2012. 

## Conclusion 

Therefore, we conducted analysis on GSS survey data and revealed a number of insights on a number of years finding patterns overtime such as religious identification change overtime, differences in religious attendance in 1973 and 2012,10 year changes in confidence in religion from 1973 to 2012, bachelor's degree attainment overtime, confidence in education overtime. We used a number of visualizetion and plots, allowing us to do exploratory data analysis and allowing us to hypothesize relationship between different variables. We conducted inference on these variables such as comparing mean education levels in US to other countries, finding confidence intervals for education levels in years 1973 and 2012 to see change in education levels, conducting inference to determine relationships between education and sex, conducting an Analysis of Variance to determine relationship between education level and race and conducting a pairwise comparison. We conducted single proportion hypothesis tests comparing bachelor's degree attainment of US to other countries, conducting independent two sample proportion hypothesis test to compare change in bachelor's degree attainment from 2002 to 2012, comparing two sample proportions hypothesis to find if bachelor's degree is different for democrats and republicans, and lastly we conduct a chu-square test to find out whether there a relationship between race and bachelor's degree attainment. Through Exploratory Data Analysis and Inference, I have answered several questions and established relationship between a number of variables in the survey. Whether correlation is causation, we will find out through regression analysis. 
